---
title: "Cleaning Data"
author: "Joseph Walker"
date: "August 9, 2017"
output: html_document
---

```{r setup, include= TRUE}
#Define Global Options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

#Load Required Packages
library(tidyverse)
library(lubridate)
library(stringr)
```
#Introduction to Cleaning Data in R

Cleaning Data is essential to the data science process. Often, it is the most labor intensive part of being a data scientist.

There are three parts:

* Exploring the raw data
* Tidying the data
* Preparing data for analysis

# Exploring Raw Data
The first part, exploring the raw data is essential for getting a feel of the data your working with and how you will need to tidy it up.

```{r checking the structure of your data}
#check the class of the dataset
class(mtcars)

#check the dimensions of the dataset
dim(mtcars)

#column names of the dataset
names(mtcars)

#check the structure of the data
str(mtcars)

#use the dplyr version of str() to view the dataset
glimpse(mtcars)

#Call a summary of the data
summary(mtcars)

#view the first few rows of the data
head(mtcars, n = 10)

#view the last rows of the data
tail(mtcars, n = 5)
```

Simple visualizations of the data can be useful

```{r data visualization exploration}
hist(mtcars$mpg)

plot(x = mtcars$disp, y = mtcars$mpg)
```

# Tidying the Data
Principles of tidy data

* Observations as rows 
* Variables as columns 
* One type of observation unit per table 

Wide vs. Long using `tidyr`

 are two functions useful for changing a wide table into long, and long to wide, respectively. These functions depend on key value pairings of the data.

`separate` and `unite`


Common Symptoms of messy data:

Headers are values, not variable names. Or vice versa. The functions `gather` and `spread` can handle these conversions by turning a dataset froma wide format to long, or long to wide, respectively. These functions depend on j=ky-value pairings. 

Multiple types of observational untis are stored in the same table. To alleviate this, you should create two separate tables that can be linked together with a primary key.

#Preparing Data For Analysis

It is often necessary to perform type conversions on your data before using it. 

To perform date conversions, the `lubridate` package is very handy. Lubridate is a part of the tidyverse but needs to be loaded explicitly.

To deal more efficiently with strings, the `stringr` package is available as part of the tidyverse.
```{r stringr functions}
#Trim leading and/or trailing white space
str_trim("   Trim the leading and trailing white space  ")
str_trim("     remove just the left side    ", side = "left")

#pad string with value/character
str_pad('21583', width = 6, side = 'left', pad = 'Y')

#str_detect

#str_replace

#tolower()

#toupper()

```

Missing & Special Values

'Inf' - Infinite

NaN - Not a Number

Use `is.na()` to detect na's in the vector.
`any(is.na())` to determine if there are any na's in the vector

`complete.cases()` will return whether or not the vector/variable has missing values.

`na.omit()` will remove any na's from the vector/variable.





