---
title: "Data Visualization with ggplot2 (Part 2)"
author: "Joseph Walker"
date: "September 26, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#load required libraries
library(tidyverse)
```

# Part 2

In Part 1, we discussed 3 elements of the grammar of graphics: data, aesthetics, & geometries. We will continue our understnading of data viz by focusing our attention on other important layers: 

* Facets 
* Statistics 
* Coordinates 
* Themes 

## Statistics

Some statistics functions and geom functions can be used synonymously in ggplot2. An example of this is the `geom_bar`, `geom_histogram` and `geom_freqpoly` functions. Under the hood, these functions are using the `stat_bin` function to plot the data. 

```{r}
#assign plot object
p <- ggplot(iris, aes(x = Sepal.Width))

#plot with geom_histogram
p + geom_histogram()

#plot with geom_bar
p + geom_bar()

#plot with stat_bin
p + stat_bin()
```

Similarly, we can apply the smoothing statistics applied by the `stat_smooth` function with `geom_smooth`.

```{r}
#assign plot object
p <- ggplot(iris, aes(Petal.Length, Sepal.Length, color = Species))

#scatter plot with least squares modeling for each individual Species, and the dataset as a whole.
#we can determine whether the confidence interval ribbon appears by setting the 'se' argument.
p + geom_point() +
  geom_smooth(method = "lm") +
  geom_smooth(method = "lm", se = FALSE, aes(group = 1))
```

