---
title: "Data Visualization"
author: "Joseph Walker"
date: "9/12/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Exploratory Visaulizations** help us explain the data.

* Helps you see what is in your data 
* Level of detail: 
    + Keep as much detail as possible 
    + Practical limit: how much can you see and interpret? 

**Explanatory Visualizations** help us share our data with others.  

* Shows others what you’ve found in your data 
* Requires editorial decisions: 
    + Highlight the key features you want to emphasize 
    + Eliminate extraneous details
   
  par(mfrow = c(rows, columns)) 
  plot, (data, type = o,l,p,s)  
    
```{r}
names(par())
# Set up a 2-by-2 plot array
par(mfrow = c(2,2))

w <- chickwts$weight
# Plot the chick weights  data as points
plot(w, type = "p")

# Add the title
title("points")

# Plot the chick weights with lines
plot(w, type = "l")

# Add the title
title("lines")

# Plot the chick weights as lines overlaid with points
plot(w, type  = "o")

# Add the title
title("overlaid")

# Plot the chick weights as steps
plot(w, type = "s")

# Add the title
title("steps")
```



```{r}
#load MASS package for the Cars93 dataset
library(MASS)

#assign datasets to variables
c1 <- Cars93
c2 <- mtcars

#compute max_hp
max_hp <- max(c1$Horsepower, c2$hp)

#compute max_mpg
max_mpg <- max(c1$MPG.city, c2$mpg)

# Create empty plot with type = "n"               
plot(max_hp, max_mpg,
     type = "n", xlim = c(0, max_hp),
     ylim = c(0, max_mpg), xlab = "Horsepower",
     ylab = "Miles per gallon")

# Add open circles to plot
points(y =c2$mpg, x =c2$hp, pch = 1)

# Add solid squares to plot
points(y = c1$MPG.city, x = c1$Horsepower, pch = 15)

# Add open triangles to plot
points(y = c1$MPG.highway, x = c1$Horsepower, pch = 2)
```


