---
title: "Data Visualization"
author: "Joseph Walker"
date: "9/12/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Exploratory Visaulizations** help us explain the data.

* Helps you see what is in your data 
* Level of detail: 
    + Keep as much detail as possible 
    + Practical limit: how much can you see and interpret? 

**Explanatory Visualizations** help us share our data with others.  

* Shows others what you've found in your data 
* Requires editorial decisions: 
    + Highlight the key features you want to emphasize 
    + Eliminate extraneous details
   
  par(mfrow = c(rows, columns)) 
  plot, (data, type = o,l,p,s)  
    
```{r}
names(par())
# Set up a 2-by-2 plot array
par(mfrow = c(2,2))

w <- chickwts$weight
# Plot the chick weights  data as points
plot(w, type = "p")

# Add the title
title("points")

# Plot the chick weights with lines
plot(w, type = "l")

# Add the title
title("lines")

# Plot the chick weights as lines overlaid with points
plot(w, type  = "o")

# Add the title
title("overlaid")

# Plot the chick weights as steps
plot(w, type = "s")

# Add the title
title("steps")
```



```{r}
#load MASS package for the Cars93 dataset
library(MASS)

#assign datasets to variables
c1 <- Cars93
c2 <- mtcars

#compute max_hp
max_hp <- max(c1$Horsepower, c2$hp)

#compute max_mpg
max_mpg <- max(c1$MPG.city, c2$mpg)

# Create empty plot with type = "n"               
plot(max_hp, max_mpg,
     type = "n", xlim = c(0, max_hp),
     ylim = c(0, max_mpg), xlab = "Horsepower",
     ylab = "Miles per gallon")

# Add open circles to plot
points(y =c2$mpg, x =c2$hp, pch = 1)

# Add solid squares to plot
points(y = c1$MPG.city, x = c1$Horsepower, pch = 15)

# Add open triangles to plot
points(y = c1$MPG.highway, x = c1$Horsepower, pch = 2)
```

Managing Visual Complexity

Too many details can detract from a plot. In addition, more plots is not necessarily more useful.

The `matplot` function allows you to plot multiple columns on an axis but as we will see, too much data can disrupt your plot.
```{r}
# Construct the vector keep_vars
keep_vars <- c("calories", "protein", "fat",
               "fibre", "carbo", "sugars")

# Use keep_vars to extract the desired subset of UScereal
df <- UScereal[, keep_vars]

# Set up a two-by-two plot array
par(mfrow = c(2,2))

# Use matplot() to generate an array of two scatterplots
matplot(y = UScereal[, c("protein", "fat")], x  = UScereal$calories, xlab = "calories", ylab = "")

# Add a title
title("Two scatterplots")

# Use matplot() to generate an array of three scatterplots
matplot(y = UScereal[, c("protein", "fat", "fibre")], x = UScereal$calories, xlab = "calories", ylab = "")

# Add a title
title(main = "Three scatterplots")

# Use matplot() to generate an array of four scatterplots
matplot(y = UScereal[, c("protein", "fat", "fibre", "carbo")], x = UScereal$calories, xlab = "calories", ylab = "")

# Add a title
title(main = "Four scatterplots")

# Use matplot() to generate an array of five scatterplots
matplot(y = UScereal[, c("protein", "fat", "fibre", "carbo", "sugars")], x = UScereal$calories, xlab = "calories", ylab = "")

# Add a title
title(main = "Five scatterplots")
```

Multiple plot arrays
```{r}
# Set up a two-by-two plot array
par(mfrow = c(2,2))

# Plot the raw duration data
plot(geyser$duration, main = "Raw data")

# Plot the normalized histogram of the duration data
truehist(geyser$duration, main = "Histogram")

# Plot the density of the duration data
plot(density(geyser$duration), main = "Density")

# Construct the normal QQ-plot of the duration data (qqPlot comes from the 'car' package)
car::qqPlot(x = geyser$duration, main = "QQ-plot")

```

Constructing & Displaing Layout Matrices

```{r}
# Use the matrix function to create a matrix with three rows and two columns
layoutMatrix <- matrix(
  c(0, 1,
    2, 0,
    0, 3), 
  byrow = TRUE, 
  nrow = 3
)

# Call the layout() function to set up the plot array
layout(layoutMatrix)

# Show where the three plots will go 
layout.show(3)
```

Now we will use a similar layout matrix and apply it to some plots using data from the `whiteside` dataset from the `MASS` package.

```{r}
# Set up the plot array
layout(layoutMatrix)

# Construct vectors indexB and indexA
indexB <- which(whiteside$Insul == "Before")
indexA <- which(whiteside$Insul == "After")

# Create plot 1 and add title
plot(whiteside$Temp[indexB], whiteside$Gas[indexB],
     ylim = c(0,8),
title("Before data only"))

# Create plot 2 and add title
plot(whiteside$Temp, whiteside$Gas,
     ylim = c(0,8),
title("Complete dataset"))

# Create plot 3 and add title
plot(whiteside$Temp[indexA], whiteside$Gas[indexA],
     ylim = c(0,8),
title("After data only"))
```

And finally, we can create arrays for different sized plots using the `layout matrix`. The `Boston` dataset from the `MASS` package us used in the following examples.

```{r}
# Create row1, row2, and layoutVector
row1 <- c(1,0,0)
row2 <- c(0,2,2)
layoutVector <- c(row1, row2, row2)

# Convert layoutVector into layoutMatrix
layoutMatrix <- matrix(layoutVector, byrow = TRUE, nrow = 3)

# Set up the plot array
layout(layoutMatrix)

# Plot scatterplot
plot(Boston$rad, Boston$zn)

# Plot sunflower plot
sunflowerplot(Boston$rad, Boston$zn)
```

