---
title: "R Notebook"
output: html_notebook
---

# Functions

Rules for writing functions:

1) Typically, if you need to copy and paste more than 2 times, it's time to write a function.
2) Solve a simple problem before writing a function.
3) A good function is both correct and understandable.

## Functional Programming


## Dealing with error

```{r safely}
library(tidyverse)
library(purrr)

df <- data.frame(x = c(1:26),
                 y = letters)
#create function (safe_log) which 'safely' takes the log of a value or vector
safe_log <- safely(log)

safe_log(10)
safe_log("a")

#variable x is all numbers so we should not see any errors
map(df$x, safe_log) %>%
  head()

# variable y contains all errors and variable z contains some errors
map (df, safe_log) %>%
  head()

safe_log(df$x)
```

## Maps over multiple arguments

```{r map over multiple arguments}
rnorm(1)
rnorm(5)
rnorm(10)

n <- c(1,5,10)

#map(.x, .f)
map(n, rnorm)


rnorm(1, mean = 1)
rnorm(5, mean = 5)
rnorm(10, mean = 10)

#map(.x, .y, .f)
map2(list(1, 5, 10), list(1, 5, 10), rnorm)

#for multiple arguments use pmap
rnorm(1, mean = 1, sd = .2)
rnorm(5, mean = 5, sd = 1)
rnorm(10, mean = 20, sd = 2.5)

pmap(list(n = list(1, 5, 10),
          mean = list(1, 5, 20),
          sd = list(.2, 1, 2.5)), rnorm)

#invoke_map iterates multiple functions over a single value
#invoke_map(list(.f), .x)
invoke_map(list(rnorm, runif, rexp), n = 5)

```

## Maps wth Side Effects

Anything that happens that isn't a return of the function  
* printing output
* plotting
* saving files

For side effects, we can use the walk() function analagous to map()
```{r}

titles <- colnames(mtcars)
pwalk(list(x = mtcars, main = titles), hist, xlab = "")

```


## Robust Functions

Functions designed for programming should be robust and strict.

3 main problems:  
* type-unstable functions
* non-standard evaluation
* hidden arguments

stopifnot function to signal errors

```{r stopifnot}

x <- 1:10

stopifnot(is.character(x))

if(!is.character(x)){
  stop(" `x` should be a character vector", call. = FALSE)
}

```


single bracket [, and sapply are examples of type inconsistent functions. In general, write functions to avoid type inconsistency.

```{r type-inconsistent}
#recall df from earlier
glimpse(df)

#create a single variable df
df2 <- data.frame(x = 1:26)

# create a function to pull the last row of a data frame
last_row <- function(x){
  x[nrow(x), ]
}

last_row(df)

str(last_row(df))

str(last_row(df2))

#to get around this type inconsistency, we can use the 'drop = FALSE' argument

last_row <- function(x){
  x[nrow(x), , drop = FALSE]}

str(last_row(df))

str(last_row(df2))

```

Non-Standard Evaluation

dplyr, ggplot


Hidden Arguments

options("digits")


